Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis

### 2018-06-08 jhir/latex-manuscript/manuscript.tex
update manuscript

\begin{abstract}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change. Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research. \hl{A significant barrier to this research is the time-consuming and resource-intensive cognitive task, the sequential analysis of pre-coded transcripts which is an integral part of the qualitative analysis of patient-counselor communication. Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating this cognitive task.} In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice. We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers. Transcripts had been previously annotated with patient-counselor behavior codes using a specialized codebook. Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients' motivational statements to guide clinical practice.

@@

\subsection{Dataset}
\label{subsec:dataset}
The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss. Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} \hl{excluding greetings, farewell and conversations for interview setup such as camera and table setup.} The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$). The experimental dataset consists of 7,192 patient, caregiver and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

@@

\section{Results}
\label{sec:results}
The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}. The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states. \hl{We observed 4,724 state transitions in the successful sequences, where 4679 and 45 state transitions occured between different and same states, respectively. On an average, 5 state transitions occurred within a sequence for both successful and unsuccessful sequences.} Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%). Successful sequences began in a state of \textit{``high receptivity''} 11\% of the time. \textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%) and affirmations (15\%). Few successful sequences began in states of \textit{``moderate receptivity''} and \textit{``low receptivity''} (2\% and 3\% of the time, respectively). These two states were characterized by different proportions of the same behaviors. \textit{``Moderate receptivity''} sequences were distinguished from \textit{``low receptivity''} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient's autonomy were about the same (14\% versus 17\%). No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%) and other-related high uptake (10\%). Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%). \textit{``Active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%). The \textit{``moderate receptivity''} \hl{state most often transitioned to \textit{``active feedback''} (27\%) and back to} \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency. The full transition matrix is presented in Table~\ref{tab:transition}.

>>>>

\begin{abstract}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change. Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research. \hl{A significant barrier to this research is the time-consuming sequential analysis of pre-coded transcripts which is an integral part of patient-counselor communication research. Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating this cognitive task.} In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice. We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers. Transcripts had been previously annotated with patient-counselor behavior codes using a specialized codebook. Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients' motivational statements to guide clinical practice.

@@

\subsection{Dataset}
\label{subsec:dataset}
The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss. Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider} \hl{excluding conversations correspond to greetings, farewell and interview setups such as table and camera settings.} The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational} and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$). The experimental dataset consists of 7,192 patient, caregiver and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

@@

\section{Results}
\label{sec:results}
The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}. The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states. \hl{We observed 4,724 state transitions in the successful sequences, where 4679 and 45 state transitions occurred between different and same states, respectively. On an average, 5 state transitions occurred within a sequence for both successful and unsuccessful sequences.} Most (84\%) successful sequences began in a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%) and affirmations (12\%). Successful sequences began in a state of \textit{``high receptivity''} 11\% of the time. \textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%) and affirmations (15\%). Few successful sequences began in states of \textit{``moderate receptivity''} and \textit{``low receptivity''} (2\% and 3\% of the time, respectively). These two states were characterized by different proportions of the same behaviors. \textit{``Moderate receptivity''} sequences were distinguished from \textit{``low receptivity''} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient's autonomy were about the same (14\% versus 17\%). No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%) and other-related high uptake (10\%). Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%). \textit{``Active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%). The \textit{``moderate receptivity''} \hl{state most often transitioned to \textit{``active feedback''} (27\%) and back to} \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency. The full transition matrix is presented in Table~\ref{tab:transition}.
